---
title: "Biomarker analysis report"
author: ""
date: '`r Sys.Date()`'
output: 
  pdf_document:
    toc: true
    toc_depth: 4
    number_sections: true
---


```{r,warning=FALSE, message=FALSE, include=FALSE}
library(gClinBiomarker)
library(knitr)

csv.name <- "/Users/lengn/Documents/Biomarker/Rpkg_Biomarkers/clone_github/gClinbiomarker_documents/Markdown_templates/example_baseline_data_2arm.csv"
outcome.class <- "survival" # this report is for two-arm survival analysis
outcome.var <- c("PFS","PFS.event") # two elements for survival endpoint
trt <- "Arm" # column name for the treatment column. this report is for two-arm survival analysis
active.code <- "TRT"
placebo.code <- "CTRL"
bm <- "KRAS.mutant"
bm.class <- "categorical"
clinical.vars <- c("Sex","Age")
clinical.vars.class <- NULL # if it is NULL, function ReadData() will learn each variable's class based on the input data frame.
input <- read.csv(csv.name, stringsAsFactors=FALSE)
# ReadData
BEP <- NULL # column name for the column which indicates biomarker evaluable population (BEP). If it is NULL, a BEP column will be generated automatically. In the generate BEP column, any entry with non missing (non-NA) biomarker value (defined by bm column) will be assigned 1.
BEP.indicator <- 1 # entries who have this label in the BEP column will be considered to be in BEP.
percentile.trycut <- NULL # no need for categorical biomarker
numerical.trycut <- NULL # no need for categorical biomarker
percentile.finalcut <- NULL # no need for categorical biomarker
numerical.finalcut <- NULL # no need for categorical biomarker
covariate <- NULL # covariate to be included in within-arm analysis, cutoff exploration and subgroup analysis. Could be a vector. If this is not NULL, an addictive model will be fitted. Estimates from the forest plot functions will be affected
strata <- NULL # stratification factors to be included in within-arm analysis, cutoff exploration and subgroup analysis. Could be a vector. If this is not NULL, a model with strata() will be fitted. Estimates from the forest plot functions will be affected

if(is.null(BEP)){
  BEP <- "BEP"
  input$BEP <- ifelse(is.na(input[[bm]]),0,1)
}
input.bep <- input[which(input[[BEP]]==BEP.indicator),]


if(is.null(clinical.vars.class)){
for(var in clinical.vars){
  if(class(input[,var])%in%c("numeric","integer"))var.class <- "numeric"
  if(class(input[,var])%in%c("logical"))class(input[,var]) <- "character"
  if(class(input[,var])%in%c("character","factor"))var.class <- "categorical"
  clinical.vars.class <- c(clinical.vars.class, var.class)
}  
}

names(clinical.vars.class) <- clinical.vars
  
if(!is.null(numerical.trycut)) percentile.trycut <- NULL
if(!is.null(numerical.finalcut)) percentile.finalcut <- NULL


stopifnot(identical(sort(unique(input[[trt]])), sort(c(active.code, placebo.code))))
input[[trt]] <- factor(as.character(input[[trt]]), levels=c(placebo.code, active.code))
```

# The dataset
The dataset have `r nrow(input)` entries. In which `r nrow(input.bep)` are in biomarker evaluable population (BEP).

* Endpoint of interest: `r outcome.var[1]`
* Biomarker: `r bm`
* Biomarker type: `r bm.class`



# Representativeness: Selection Bias of Biomarker Population

In this section, we are trying to answer the question: *Are biomarker evaluable population representative of the full population population? *

Key baseline demographics and prognostic characteristics (including stratification variables and any variables with known prognostic effect) and efficacy outcomes should be summarized by treatment groups and compared between biomarker evaluable population (BEP) and the full population. These analyses are conducted to investigate any potential selection bias associated with the availability of the biomarker (e.g. we may not get enough tissue for patients whose tumor size is small. Therefore they may be exlcuded from BEP).

The key clinical biomarkers considered are:

```{r}
clinical.vars.class
```



## Check selection bias in terms of key clinical variables, between full population and BEP


```{r}
kable(SummaryVars(data=input, trt=trt, subgroup=BEP, subgroup.indicator = BEP.indicator, var=clinical.vars, var.class=clinical.vars.class))
```


## Check whether the clinical outcome in BEP is comparable to the full population

The following plot compares survival outcome in BEP vs. the full population. The KM curve and 95% CI are plotted for each arm. The BEP KM curve is expected to be within the full population confidence bands.

```{r, fig.height=5, fig.width=10}
CompareKM(data=input,tte=outcome.var[1], cen=outcome.var[2],trt=trt, bep=BEP, bep.indicator = BEP.indicator)
```



## Examine whether the prognostic/predictive/null trend of key clinical variables holds in BEP

The following forest plot can be used to examine whether any of the key prognostic/predictive clinical variables still show prognostic/predictive trend in BEP:

```{r, fig.height=14,fig.width=10}
forest.bep <- PlotTabForestMulti(data=input,
                  outcome.class=outcome.class,
                  outcome.var=outcome.var,
                  trt=trt,
                  var=clinical.vars,
                  var.class=clinical.vars.class,
                  bep=BEP,bep.indicator=BEP.indicator, 
                  compare.bep.itt=TRUE
                   )
```




## Compare treatment effect estimation in full population and in BEP, adjusted for key clinical variables
The following analyses show summary statistic to look at  the trt/ctrl (target/reference) HR in full population and trt/ctrl HR in BEP. Both unadjusted and adjusted analyses are performed

```{r}


kable(
  CoxTab(data=input, tte=outcome.var[1], cens=outcome.var[2], var=trt, 
       var.class="categorical"),
caption="full population, unadjusted"
)

kable(
  CoxTab(data=input, tte=outcome.var[1], cens=outcome.var[2], var=c(trt, clinical.vars), var.class=c("categorical",clinical.vars.class)),
caption="full population, adjusted for clinical variables"
)

kable(
  CoxTab(data=input.bep, tte=outcome.var[1], cens=outcome.var[2], var=trt, 
       var.class="categorical"),
caption="BEP, unadjusted"
)
kable(
  CoxTab(data=input.bep, tte=outcome.var[1], cens=outcome.var[2], var=c(trt, clinical.vars), var.class=c("categorical",clinical.vars.class)),
caption="BEP, adjusted for clinical variables"
)

```
  


If any selection bias is suspected, you may consider to stratify for the imbalanced factor in downstream analysis (e.g. unstratified analysis as primary analysis and stratified analysis as sensitivity analysis).


# Biomarker property and its association to clinical variables

Relationship between the biomarker and key demographic and prognostic variables should also be investigated using bivariate plots. Prognostic property of the biomarker should also be assessed, by estimates of the clinical efficacy in the control arm.


## Biomarker property and relationship to clinical variable

The following results show single variate plot for biomaker and bi-variate plots to investigate biomarker-clinical variable relationship.

```{r fig.width=12, fig.height=8}
PlotProperty(data=input, biomarker.var=bm, biomarker.class=bm.class,
             var=clinical.vars, 
             var.class=clinical.vars.class,
             log2=FALSE, par.param = list(mfrow=c(2,3)))
```

## Whether the biomarker shows within-arm effect

The following figures investigate whether the biomarker shows a within-arm effect (e.g. patients within one biomarker subgroup tend to have better clinical outcome):
```{r fig.width=8, fig.height=4}
input.ctrl <- subset(input, Arm==placebo.code) ## Data with only ctrl samples
res.multicut.ctrl <- PlotTabForestBiomarker(data=input.ctrl,
                                  outcome.class=outcome.class,
                                  outcome.var=outcome.var,
                                  var=bm, 
                                  var.class=bm.class,
                                  percentile.cutoff=percentile.trycut, 
                                  numerical.cutoff = numerical.trycut,
                                  main.prefix=placebo.code,
                                  greater=TRUE, less=FALSE,
                                  covariate=covariate, strata=strata)

input.trt <- subset(input, Arm==active.code) ## Data with only ctrl samples
res.multicut.trt <- PlotTabForestBiomarker(data=input.trt,
                                  outcome.class=outcome.class,
                                  outcome.var=outcome.var,
                                  var=bm, 
                                  var.class=bm.class,
                                  percentile.cutoff=percentile.trycut, 
                                  numerical.cutoff = numerical.trycut,
                                  main.prefix=active.code,
                                  greater=TRUE, less=FALSE,
                                  covariate=covariate, strata=strata)
```

The forest plots above show within-arm HR across biomarker subgroups.
For a given arm, if the HR is not all around 1, it indicates that within this arm the biomarker has an association to the clinical outcome.


If similar trend is seen in both arms, it indicates that the biomarker may have a prognostic effect (the biomarker is able to identify patients with better/worse clinical outcome, regardless of treatment).




# Biomarker subgroup analysis


## Estimations within each subgroup

The following figure shows estimate of treatment effect in biomarker subgroups:

```{r, fig.height=5,fig.width=8 }
if(bm.class=="numeric"){
if(!is.null(numerical.finalcut)) levs <- paste0(c(">=","<"),numerical.finalcut)
if(is.null(numerical.finalcut)) {
  nm <- quantile(input.bep[[bm]],percentile.finalcut, 2) # default quantile type in forest plot functions
  numerical.finalcut <- round(nm,2) # default rounding decimal in forest plots
  levs <- paste0(c(">=","<"),percentile.finalcut*100,"%")
}

bm2 <- paste0(bm,"_group")
input[[bm2]] <- ifelse(input[[bm]] >= numerical.finalcut, levs[1],levs[2])
input[[bm2]]<- factor(input[[bm2]], levels=levs) # ">=" as Dx+
}

if(bm.class=="categorical") {
  bm2 <- bm
  levs <- unique(input[[bm2]])
}

res.2group <- PlotTabForestBiomarker(data=input,
                                  outcome.class=outcome.class,
                                  outcome.var=outcome.var,
                                  trt=trt,
                                  var=bm2,
                                  var.class="categorical",
                                  greater=TRUE, less=TRUE,
                                  show.itt=TRUE, show.bep=TRUE,
                                  covariate=covariate, strata=strata)
```



## KM curves

The following figure show KM curves of the biomarker subgroups:

```{r  include=TRUE,fig.width=9, fig.height=9}
km.out <- PlotKM(data=input, tte=outcome.var[1],cen=outcome.var[2], bep=BEP,  
        trt=trt, var=bm2, var.class="categorical",
        legend.loc="topright",
        plot.median=FALSE)
```


## Check whether biomarker subgroup is confounded with key clinical variables

The following table checks whether the biomarker subgroup is confounded with clinical variables. Here we check whether the clinical variable distribution is comparable in biomarker subgroups. 

\tiny
```{r}
input.bep <- input[which(input[[BEP]]==BEP.indicator),]


kable(
SummaryVars(data=input.bep,trt=trt, subgroup=bm2, var=clinical.vars, 
       var.class=clinical.vars.class, subgroup.indicator=levs[1],compare.subgroup=TRUE)
)
```

\normalsize

The following plot show treatment effect estimations in smaller subgroups defined by both biomarker and clinical variables. For numerical clinical variable, it is dichotomized by its median.

```{r, fig.height=7,fig.width=10}
res.subgroup.cov <- PlotTabForestMulti(data=input,
                                  outcome.class=outcome.class,
                                  outcome.var=outcome.var,
                                  trt=trt,
                                  var=clinical.vars,
                                  var.class=clinical.vars.class,
                                  compare.bep.itt=FALSE,
                                  compare.subgroup=TRUE,
                                  subgroup=bm2,
                                  covariate=covariate, strata=strata
                   )
```


